### 2013-08-23 doc/dissertation/document/backmatter/extra_scaling.tex
finished parallel corrections

\chapter{Subdomain Neumann-Ulam Scaling with Multiple Sets}
\label{chap:extra_scaling}
As an additional analysis of MCSA using subdomain Monte Carlo, the multiple
sets scaling studies are repeated to determine if any performance
benefits are made from replication by eliminating Monte Carlo
communication. Figure~\ref{fig:titan_strong_subdomain_ms_time} gives
the runtimes for the strong scaling calculation and
Figure~\ref{fig:titan_strong_subdomain_ms} the computed absolute
efficiencies. For the weak scaling studies,
Figure~\ref{fig:titan_weak_subdomain_ms_time} gives the wall times and
Figure~\ref{fig:titan_weak_subdomain_ms} the efficiencies. For both
scaling studies, the results are effectively the same as for the case
with Monte Carlo communication. This is again due to the fact that
adding sets to the problem is not only an exercise in strong scaling
but it also adds additional parallel overhead in the form of the tally
vector reduction across blocks.

From a parallel performance perspective, MCSA may then be best used in
a mode where the Monte Carlo is performed over only the subdomains
with no replication or overlap added to the system. In this mode,
iterative performance is maintained relative to the solutions with
communication even for more ill-conditioned problems with larger
spectral radii. Furthermore, no overlap was required to maintain
iterative performance be letting histories near the boundary take a
few more steps in their random walks. If in the future there are
ill-conditioned problems where using subdomain Monte Carlo reduces the
algorithmic efficiency of MCSA due to the information lost in the
correction vector, overlap would be a viable mechanism to potentially
maintain that iterative performance, keeping in mind the potential
impacts on scalability that were observed in this work. However,
although eliminating all Monte Carlo communication was observed to be
the best mode of use for MCSA, the parallel Monte Carlo methods
developed for this work still provide value for cases where only a
Monte Carlo solution is desired. In those cases, full transport of the
Monte Carlo histories are required in order to generate the correct
solution.

>>>>

\chapter{Subdomain Neumann-Ulam Scaling with Multiple Sets\ }
\label{chap:extra_scaling}
As an additional analysis of MCSA using subdomain Monte Carlo, the multiple
sets scaling studies are repeated to determine if any performance
benefits are made from replication by eliminating Monte Carlo
communication. Figure~\ref{fig:titan_strong_subdomain_ms_time} gives
the runtimes for the strong scaling calculation and
Figure~\ref{fig:titan_strong_subdomain_ms} the computed absolute
efficiencies. For the weak scaling studies,
Figure~\ref{fig:titan_weak_subdomain_ms_time} gives the wall times and
Figure~\ref{fig:titan_weak_subdomain_ms} the efficiencies. For both
scaling studies, the results are effectively the same as for the case
with Monte Carlo communication. This is again due to the fact that
adding sets to the problem is not only an exercise in strong scaling
but it also adds additional parallel overhead in the form of the tally
vector reduction across blocks.

### 2013-08-23 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
finished parallel corrections

Using the Neumann-Ulam solver to compute the MCSA correction only in
the subdomains, the strong and weak scaling studies were repeated to
assess the performance gains made by eliminating communication in the
Monte Carlo solver. For each scaling study, the 16-core base case with
Monte Carlo communication was used as the reference point for
efficiency computations in order to measure how eliminating
communication boosts
efficiencies. Figure~\ref{fig:titan_strong_subdomain} gives the
results of the strong scaling study and
Figure~\ref{fig:titan_weak_subdomain} gives the results of the the
weak scaling study in terms of absolute efficiencies. In both cases,
the benefits to parallel performance of using MCSA as a subdomain
solver are obvious. In the strong scaling case we make the important
observation that we hit the strong scaling wall at the same time with
and without the additional Monte Carlo communication. Although the
Monte Carlo communication did contribute to the efficiency decrease,
it is clear that the remaining matrix and vector operations are the
primary impediment to scalability. Furthermore, we see an even greater
rise above unity in efficiencies because there is less communication
overhead hiding the better cache utilization as the local problem size
shrinks. At 16,384 cores, performing the subdomain solves boost the
absolute efficiency from 38\% to 67\% relative to the reference case.

In the weak scaling study, we do not have the scaling wall to contend
with and eliminating Monte Carlo communication through subdomain
solves provides an effectively linear boost in the weak scaling
absolute efficiency. We still see a small reduction in weak scaling
efficiency as a function of core count due to the communication
overhead for the parallel matrix and vector operations. We observe
super unitary efficiencies for the subdomain case as there was a
substantial improvement in runtimes relative to the 16-core base case
with Monte Carlo communication.

@@

Of additional concern for the subdomain method is the potential
decrease in iterative performance due to the fact that histories near
the boundary will potentially have their random walk truncated, thus
decreasing the information available to compute the correction
vector. For both the weak and strong scaling study, all calculations
with split histories converged in 22 or in only a few cases 21
iterations for both the subdomain solve and the solve with
communication giving an algorithmic efficiency of 1.048 for all
cases. This small improvement in iteration count was not observed to
be a function of problem size, number of processors, or whether the
subdomain method was used. For the replicated history cases, the
algorithmic efficiency improvement was not as good as when
communication was allowed for the Monte Carlo algorithm. For these
calculations, the number of iterations to converge was reduced to 18
or 19 in both the strong and weak scaling studies with lower core
counts converging in fewer iterations.

Therefore, there is no reason to expect overlap to improve the
iterative or scaling performance of the method as iterative
performance is maintained without it for this particular
problem. Overlap is not as effective because histories relinquish most
of their useful information within the first few steps (which is why
we can use such a large weight cutoff and there is an insensitivity to
weight cutoff as shown in
Figure~\ref{fig:estimator_wc_iters}). Keeping them around longer with
overlap to do extra work in the local domain will not improve
iterative performance and thus increase compute times. Supplemenatary
scaling studies using the subdomain Neumann-Ulam method with multiple
sets are presented in Appendix~\ref{chap:extra_scaling}.

@@

\begin{itemize}
\item MCSA can be effectively parallelized using domain decomposition
\item The leakage fraction of stochastic histories from a parallel
  subdomain can be quantified analytically using algebraic quantities
  (see Appendix~\ref{chap:parallel_theory})
\item The multiple-set overlapping-domain (MSOD) parallel algorithm
  for particle transport has been adapted to parallelize MCSA
\item MCSA parallelized with MSOD was verified to produce the same
  results for solutions to the neutron diffusion equation as
  production Krylov methods
\item Strong scaling studies were performed with up to 65,536 cores
  and weak scaling studies were performed with up to 131,072 cores on
  the Titan Cray XK7 leadership class machine at Oak Ridge National
  Laboratory
\item MCSA scales favorably compared to production Krylov methods for
  both strong and weak scaling cases
\item Overlap in small quantities determined analytically by random
  walk length can provide parallel efficiency boosts of a few percent
  in strong scaling cases but is ineffective in weak scaling cases
\item Multiple sets will never boost parallel efficiencies in the weak
  scaling case due to additional communication overhead from added
  reduction operations but will boost efficiencies beyond the strong
  scaling wall
\item Multiple sets offers a means to reduce time to solution by
  solving multiple copies of the original problem and combining the
  solutions using superposition
\item MCSA is most efficiently used in parallel as a stochastic
  realization of an additive Schwarz method
\item With future optimization, large scale reactor analyses in which
  the transport operator is fully formed (e.g. the $SP_N$ equations)
  could experience improved time to solution on leadership-class
  machines with MCSA given the improved scaling performance
\end{itemize}

>>>>

Using the Neumann-Ulam solver to compute the MCSA correction only in
the subdomains, the strong and weak scaling studies were repeated to
assess the performance gains made by eliminating communication in the
Monte Carlo solver. For each scaling study, the 16-core base case with
Monte Carlo communication was used as the reference point for
efficiency computations in order to measure how eliminating
communication boosts
efficiencies. Figure~\ref{fig:titan_strong_subdomain} gives the
results of the strong scaling study and
Figure~\ref{fig:titan_weak_subdomain} gives the results of the the
weak scaling study in terms of absolute efficiencies.

In the strong scaling case we make the important observation that we
hit the strong scaling wall at the same time with and without the
additional Monte Carlo communication. Although the Monte Carlo
communication did contribute to the efficiency decrease, it is clear
that the remaining matrix and vector operations are the primary
impediment to scalability. Furthermore, we see an even greater rise
above unity in efficiencies because there is less communication
overhead hiding the better cache utilization as the local problem size
shrinks. At 16,384 cores, performing the subdomain solves boost the
absolute efficiency from 38\% to 67\% relative to the reference case.

In the weak scaling study, we do not have the scaling wall to contend
with and eliminating Monte Carlo communication through subdomain
solves provides an effectively constant boost in the weak scaling
absolute efficiency. We still see a small reduction in weak scaling
efficiency as a function of core count due to the communication
overhead for the parallel matrix and vector operations. We observe
super unitary efficiencies for the subdomain case as there was a
substantial improvement in runtimes relative to the 16-core base case
with Monte Carlo communication.

@@

From a parallel performance perspective, MCSA may then be best used in
a mode where the Monte Carlo is performed over only the subdomains
with no replication or overlap added to the system. In this mode,
iterative performance is maintained relative to the solutions with
communication even for more ill-conditioned problems with larger
spectral radii. If in the future there are ill-conditioned problems
where using subdomain Monte Carlo reduces the algorithmic efficiency
of MCSA due to the information lost in the correction vector, overlap
would be a viable mechanism to potentially maintain that iterative
performance, keeping in mind the potential impacts on scalability that
were observed in this work. However, although eliminating all Monte
Carlo communication was observed to be the best mode of use for MCSA,
the parallel Monte Carlo methods developed in this work still provide
value for cases where only a Monte Carlo solution is desired. In those
cases, full transport of the Monte Carlo histories are required in
order to generate the correct solution.

@@

\begin{itemize}
\item MCSA can be effectively parallelized using domain decomposition
\item The leakage fraction of stochastic histories from a parallel
  subdomain can be quantified analytically using algebraic quantities
  (see Appendix~\ref{chap:parallel_theory})
\item The multiple-set overlapping-domain (MSOD) parallel algorithm
  for particle transport has been adapted to parallelize MCSA
\item MCSA parallelized with MSOD was verified to produce the same
  results for solutions to the neutron diffusion equation as
  production Krylov methods
\item Strong scaling studies were performed with up to 65,536 cores
  and weak scaling studies were performed with up to 131,072 cores on
  the Titan Cray XK7 leadership class machine at Oak Ridge National
  Laboratory
\item MCSA scales favorably compared to production Krylov methods for
  both strong and weak scaling cases
\item Overlap in small quantities determined analytically by random
  walk length can provide parallel efficiency boosts of a few percent
  in strong scaling cases but is ineffective in weak scaling cases
\item Multiple sets will never boost parallel efficiencies in the weak
  scaling case due to additional communication overhead from added
  reduction operations but will boost efficiencies beyond the strong
  scaling wall
\item Multiple sets offers a means to reduce time to solution by
  solving multiple copies of the original problem and combining the
  solutions using superposition
\item MCSA is most efficiently used in parallel as a stochastic
  realization of an additive Schwarz method
\end{itemize}
